---
title: "Class 10: Halloween mini project"
author: "Serena Quezada (PID: A18556865)"
format: pdf
toc: true
---

As it is nearly Halloween and the half way point in the quarter let's do a mini project to help us figure out the best candy!

Our come from the 538 website and is available as a CSV file:

## Data Import 

```{r}
candy <- read.csv("Candy data.csv", row.names = 1)
head(candy)
```
```{r}
flextable::flextable(head(candy, 10))
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```


## 2. What is your favorite candy?

```{r}
candy["Twix", ]$winpercent
```

> Q3. What is your favorite candy in the dataset and what is it's winpercent value?

My favorite candy is Twix and its winpercent value is 81.64%

```{r}
candy["Twix", ]$winpercent
```

With tidyverse!

```{r}
library(dplyr)

candy |>
  filter(rownames(candy)=="Twix") |>
  select(winpercent)
```

> Q4. What is the winpercent value for "Kit Kat"?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

## Quick overview of the dataset

```{r}
library("skimr")
skim(candy)

```
> Q6.  Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent is on a 0-100 scale the rest are 0-1 scale 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

That the candy does not contain chocolate 

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent) +
  geom_histogram(bins = 20)
```
> Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy) + 
  aes(winpercent) +
  geom_density(bins = 20)
```

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```

The mean is above 50% and the median is below 50%, it's in between 47% - 50% so it's below 50%. 

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy in the dataset
# 2. Find their winpercent values 
# 3. Calculate the mean of these values 

# 4-6. Do the same for fruity candy
# 7. Compare mean winpercents of chocolate vs fruity
# 8. Pick the highest as the winner

```

Steps # 1-3:
```{r}
choc.inds <- candy$chocolate==1
choc.win <- candy[choc.inds, ]$winpercent
choc.mean <- mean(choc.win)
choc.mean
```

Steps # 4-6: 
```{r}
fruity.inds <- candy$fruity==1
fruity.win <- candy[fruity.inds, ]$winpercent
fruity.mean <- mean(fruity.win)
fruity.mean
```

Chocolate candy has a higher winpercent compared to fruity candy. 

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruity.win)
```
This test is statistically significant 

## 3. Overall Candy Rankings 

> Q13. What are the five least liked candy types in this set?

```{r}
candy |>
  arrange(winpercent) |>
  head(5)
```

```{r}
x <- c(5,1,10,4)
#sort(x)
order(x)
#(candy$winpercent)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(winpercent) |>
  tail(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values 

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
HINT: you can use aes(winpercent, reorder(rownmaes(candy), winpercent)) to improve your plot

```{r}
ggplot(candy) +
  aes(x = winpercent, 
      y = reorder(rownames(candy), winpercent)) +
  geom_col()
```
Adding colors based on the "type of candy"

```{r}

my_cols<-rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] <-  "chocolate"
my_cols[as.logical(candy$bar)] <-  "brown"
my_cols[as.logical(candy$fruity)] <-  "pink"

ggplot(candy) +
  aes(x = winpercent, 
      y = reorder(rownames(candy), winpercent)) +
  geom_col(fill = my_cols)
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is sixlets

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is starbust

## 4. Winpercent and Pricepercent

A plot with both variables/columns winpercent and pricepercent 

```{r}
library(ggrepel)

ggplot(candy) +
  aes(x = winpercent,
      y = pricepercent, 
      label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(max.overlaps = 7)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

The highest ranked candy in terms of winpercent for the least money is chocolate

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
tail( candy[ord,c(11,12)], n=5 )
```

## 5. Exploring the correlation structure 

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The two variables that are anti-correlated are chocolate and fruity candy

> Q23. Similarly, what two variables are most positively correlated?

The two variables that are positively correlated are chocolate and bar.  

## 6. Principal Component Analysis

The function to use is called `prcomp()` with an optional `scale=T/F` argument. 

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```
Our main PCA result figure

```{r}
plot(pca$x [, 1:2])
```
Add some color!

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
We should also examine the variable "loadings" or contributions of the original variables to the new PCs 

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```
> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? 

The original variables that are picked up strongly by PC1 in the positive direction are hard, fruity and pluribus candy. This makes sense to me because fruity is a highly favored type of candy. 

```{r}
p <- ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) +
  geom_point(col = my_cols) +
  geom_text_repel(col = my_cols)
```

Interactive plots that can be zoomed on and "brushed" over can be made with the **plotly** package. It's output is interactive and will not render to PDF

```{r}
library(plotly)

#plotly(p)
```


